Hydras \& Co.

### 2023-11-02 doc/chapter-primrec.tex
New scopes (#167)
* UTF8 symbols for beta function

* experiment custom entries for fol propositions

* More experiments with notations (unstable)

* New notations for PR functions (to complete)

* New notations for PR functions and FOL formulas (experimental)
Minor changes (mainly notations)

Primitive recursive functions are total functions from $\mathbb{N}^n$ to
$\mathbb{N}$, for some $n\in\mathbb{N}$. Note that not all 
total $n$-ary recursive functions are primitive recursive
(see for instance Sect.~\vref{sect:ack-not-PR}).

@@

For instance, the projection $\pi{2,3}$ satisfies the equation
$\pi{1,3}(x,y,z)=y$ for any natural numbers $x$, $y$ and $z$.

@@

\subsection{Addition on natural natural numbers}

@@

\begin{align}
  0 + x_1 &:= x_1 \\
  S x_0 + x_1 &:= S (x_0 + x_1)
\end{align}

@@

\begin{align}
0 + x_1 &:= g(x_1) \\
S p + x_1 &:= h(p, p+x_1, x_1) \\
\textbf{where}\; g(x_1)&:= x_1 \\
\textbf{and}\; h(p,x, x_1) &:= S\;x 
\end{align}

@@

\subsection{Multiplication on natural natural numbers}

@@

\begin{align}
  0 \times x_1 &:= 0 \\
  S x_0 \times x_1 &:= (x_0 \times x_1) + x_1
\end{align}

This function is an instance of the primitive recursion scheme,
with $n=1$, $g$ the constant unary function which returns $0$ (see Subsection~\vref{sect:k0}),
 and $h$ the function defined by $h(p,x,n_1)=x+n_1$, which can be written as the composition of $+$ and the projections 
$\pi_{2,3}$ and $\pi_{3,3}$ (The three of them  being  primitive recursive).

\index{hydras}{Exercises}
\begin{exercise}
 Show that the factorial function is primitive recursive (by the same method as above). 
\end{exercise}

@@

The four examples above are defined in \coq in Section~\vref{sect:primrec-examples2}

@@

O'Connor's library on Gödel's incompleteness theorems contains a little more 
than 45K lines of scripts. The part dedicated to primitive recursive functions and Peano arithmetic is 32K lines long and is originally structured in 38 modules.
Thus, we propose a partial exploration of this library, through examples and exercises. Our additions to the original library --- mainly examples and counter-examples ---,
are stored in the directory \texttt{theories/ordinals/MoreAck}.

@@

@@

@@

\input{movies/snippets/PrimRecExamples/evalPrimRecEx}

@@

\subsubsection{Constant function of value 0}

\inputsnippets{PrimRecExamples/cst0}

\subsubsection{Addition}

\inputsnippets{PrimRecExamples/addition}

\subsubsection{Multiplication}

\inputsnippets{PrimRecExamples/multiplication}

\subsubsection{The factorial function}

\inputsnippets{PrimRecExamples/factorial}

@@

\begin{align}
A(0,n)&=n+1\\
A(m+1,0)&=A(m,1)\\
A(m+1,n+1)&=A(m,A(m+1,n))
\end{align}

>>>>

\emph{Primitive recursive functions} are total functions from $\mathbb{N}^n$ to
$\mathbb{N}$, for some $n\in\mathbb{N}$. Note that not all 
total $n$-ary recursive functions are primitive recursive
(see for instance Sect.~\vref{sect:ack-not-PR}).

Primitive recursive \emph{relations} are boolean total functions whose \emph{characteristic function} --- obtained by mapping the returned value to $1$ (\texttt{true}) or $0$ (\texttt{false}) ---  is primitive recursive.

\section{Mathematical definition}
\label{sect:primrecmathdef}

@@

For instance, the projection $\pi_{2,3}$ satisfies the equation
$\pi_{2,3}(x,y,z)=y$ for any natural numbers $x$, $y$ and $z$.

@@

\subsubsection{Addition on natural natural numbers}

@@

\begin{align*}
  0 + x_1 &:= x_1 \\
  S x_0 + x_1 &:= S (x_0 + x_1)
\end{align*}

@@

\begin{align*}
&0 + x_1 := g(x_1) \\
& S \; p +x_1 := h(p, p+x_1, x_1) \\
&\textbf{where}\; g(x_1):= x_1 \\
&\textbf{and}\; h(p,x, x_1) := S\;x 
\end{align*}

@@

\subsubsection{Multiplication on natural natural numbers}

@@

\begin{align*}
  0 \times x_1 &:= 0 \\
  S x_0 \times x_1 &:= (x_0 \times x_1) + x_1
\end{align*}

This function is an instance of the primitive recursion scheme,
with $n=1$, $g$ the constant unary function which returns $0$ (see subsection~\vref{sect:k0}),
 and $h$ the function defined by $h(p,x,n_1)=x+n_1$, which can be written as the composition of $+$ and the projections 
$\pi_{2,3}$ and $\pi_{3,3}$ (the three of them  being  primitive recursive).

\index{hydras}{Exercises}
\begin{exercise}
 Build a primitive recursive definition of the factorial function,
using the constructions of  Section~\ref{sect:primrecmathdef},
addition and multiplication.
\end{exercise}

@@

The four examples above are defined as \coq terms in Section~\vref{sect:primrec-examples2}

@@

O'Connor's library on Gödel's incompleteness theorems contains a little more 
than 45K lines of code. The part dedicated to primitive recursive functions and Peano arithmetic is 32K lines long and is originally structured in 38 modules.
Thus, we propose a partial exploration of this library, through examples and exercises. Our additions to the original library --- mainly examples and counter-examples ---,
are stored in the directory \texttt{theories/ordinals/MoreAck}.

@@

In order to make the terms of type \texttt{Primrec $n$} more readable, we introduce some notations, mainly inspired by \coq's standard library's notations for vectors.

\inputsnippets{primRec/PRNotations}

\noindent
\emph{From \href{../theories/html/hydras.MoreAck.PrimRecExamples.html}{MoreAck.primRecExamples}.}

\inputsnippets{PrimRecExamples/FirstExamples}

@@

\subsubsection{A few tests}
Let us consider again the functions defined in 
previous sections.

@@

\input{movies/snippets/PrimRecExamples/tests}

@@

\subsubsection{A few proofs of correctness}

It is tempting to \emph{prove} that our functions \texttt{cst0},
\texttt{plus}, \texttt{mult} and \texttt{fact} are correct implementations in \texttt{PrimRec} of the mathematical functions we
consider.

\input{movies/snippets/PrimRecExamples/correctness}

@@

\begin{align*}
A(0,n)&=n+1\\
A(m+1,0)&=A(m,1)\\
A(m+1,n+1)&=A(m,A(m+1,n))
\end{align*}
